---
title: "DATA607_Assignment6"
author: "Gabriel Santos"
date: "10/01/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Assignment – Tidying and Transforming Data

The chart above describes arrival delays for two airlines across five destinations.  \

Your task is to:

(1) Create a .CSV file  that includes all of the information above.  \
    You’re encouraged to use a “wide” structure similar to how the information appears above, \ 
    so that you can practice tidying and transformations as described below. \
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as \ 
   needed to tidy and transform your data.\   
(3) Perform analysis to compare the arrival delays for the two airlines.\ 
(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative descriptions of your data\     cleanup work, analysis, and conclusions.\
Please include in your homework submission: \
The URL to the .Rmd file in your GitHub repository. and 
The URL for your rpubs.com web page

### Prerequisites

Load required packages

```{r message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
```

### Read the airline data


### First read

```{r}
#reading .csv file
airlines = read.csv("https://raw.githubusercontent.com/GabrielSantos33/DATA607-Assignment-6/main/airlines.csv")
#eliminating empty row
airlines=airlines[!(rowSums(is.na(airlines))),]
#Populating empty airline cells
airlines[2,1] = "ALASKA"
airlines[4,1] = "AM WEST"
#tidying things up
airlines1 = gather(airlines, city, freq, Los.Angeles:Seattle)
airlines2 = spread(airlines1, X.1, freq)
names(airlines2)[1] = 'airlines'
tbl_df(airlines2)
```

### Second read after modifying the csv file.

```{r message=FALSE}
airlinesUntidyDf<-read.csv("https://raw.githubusercontent.com/GabrielSantos33/DATA607-Assignment-6/main/airlines2.csv")
airlinesUntidyDf
```


### Clean up the data

*Transform the data from wide format to long format*

```{r message=FALSE}
#Move the destination columns (3 to 7) from the untidy data frame  into single destination column
#wide data into longer format and then widen the Arrival Status into separate columns
airlinesDf <- gather(airlinesUntidyDf,Destination, delays, 3:7) %>%spread(2,4)
airlinesDf
```


### Data transformation

*Add additional columns for total ,  percentages for On time and delays*

```{r message=FALSE}
#Add additional columns for total flights , on time and delayed percentages 
airlinesTfDf <- airlinesDf %>% 
        mutate(total = delayed + `on time`,prcntOnTime = round(100 * `on time` / total, 2), prcntDelayed = round(delayed / total * 100, 2)) 
airlinesTfDf
#summary(airlinesTfDf)
#Create a summary record for each airline
 airlineSummaryDf <- airlinesTfDf %>% 
  group_by(Airline) %>% 
  summarise(totalDelayed = sum(delayed), totalOnTime = sum(`on time`), totalFlights = sum(total)) 
 
# Add additional columns for percentage - on-time and delayed flights 
 airlineSummaryDf<- airlineSummaryDf %>%  mutate(percentOntime = round(totalOnTime/totalFlights * 100, 2),percentDelayed = round(totalDelayed/totalFlights * 100, 2))
airlineSummaryDf
```



### Data visualisation

*Bar plot for each destination by Airline*

```{r message=FALSE}
#bar plot for ontime flights
plt <- ggplot(airlinesTfDf, aes(y=prcntOnTime , x = Airline, color = Airline, fill = Airline)) + 
    geom_bar( stat = "identity") +
    geom_text(aes(label = prcntOnTime ), vjust = -.35) +
     ylim(0, 100) +
    facet_wrap(~Destination) +
    ylab("Percentage - On time Flights") +
    ggtitle("Percentage of On time Flights / Airline / Destination")
    
plt
# bar plot for delayed flights
plt1 <- ggplot(airlinesTfDf, aes(y=prcntDelayed, x = Airline, color = Airline, fill = Airline)) + 
    geom_bar( stat = "identity") +
    geom_text(aes(label = prcntDelayed), vjust = -.35) +
    ylim(0, 35) +
    facet_wrap(~Destination) +
    ylab("Percentage - Delayed Flights") +
    ggtitle("Percentage of Delayed Flights / Airline / Destination")
    
plt1
```


### Summary

*According to the graphs of each airline by destination, the ALASKA airline has a higher percentage of flight departure on time than the AM WEST airline, in the different destinations.*

*The AM WEST airline has a higher percentage of delayed flights than ALASKA, in the different destinations.*

*In the general analysis, considering the total flights, the ALASKA airline has a higher percentage of delayed flights with 13.3% and the AM WEST airline has 10.69% of flight delays and this is because the ALASKA airline has a total of 3775 flights while the AM WEST airline has a total of 7225 flights.*

*In conclusion, the AM WEST airline has a higher percentage of departure on time of 89.1% compared to ALASKA airlines.*
